---
title: "0. climattR — Package overview & full workflow"
author: "Marc Lemus-Canovas"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{0. climattR — Package overview & full workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center"
)
```

## What is climattR?

**climattR** is an R package for **rapid attribution of extreme weather events to
climate change** using the *flow analogues* methodology. Given a target event (e.g.
a heatwave or storm), the package:

1. Identifies historical days with the same large-scale atmospheric pattern (Z500)
2. Reconstructs the surface conditions (temperature, precipitation) those days produced
3. Compares conditions across two climate sub-periods (counterfactual past vs. factual present)
4. Quantifies the attribution signal: Probability Ratio (PR), Fraction Attributable Risk (FAR), $\Delta T$

The approach is rapid (minutes on a laptop), fully reproducible with open reanalysis
data (ERA5), and methodologically consistent with the World Weather Attribution
(WWA) framework.

---

## Workflow diagram

```{r workflow-diagram, echo=FALSE, fig.height=7, fig.width=9}
library(ggplot2)

# --- node definitions ---
nodes <- data.frame(
  id    = 1:7,
  label = c(
    "1. prepare_data()\nNetCDF → SpatRaster\ntemporal window + detrend",
    "2. analogs_searcher()\nRMSD / Euclidean / Pearson\ntop-N analog days",
    "3a. bs_spatanalogs()\nBootstrap spatial\nreconstruction (SpatRaster)",
    "3b. as_ts() +\nbs_analogs()\nBootstrap tabular\n(climattr_bs)",
    "4. attribution_metrics()\nPR · FAR · ΔT\n95% CI",
    "5. autoplot()\nDensity: factual vs.\ncounterfactual",
    "6. downscale_temperature()\nKriging + DEM\nhigh-res grid"
  ),
  x     = c(4, 4, 2, 6, 6, 4, 2),
  y     = c(9, 7, 5, 5, 3, 3, 3),
  type  = c("input", "core", "core", "core", "output", "output", "optional")
)

# colour palette
pal <- c(input = "#4575b4", core = "#1a9850", output = "#d73027", optional = "#f4a582")

# --- arrow definitions ---
arrows <- data.frame(
  x    = c(4, 4, 4, 6, 6, 2),
  y    = c(8.55, 6.55, 6.55, 4.55, 4.55, 4.55),
  xend = c(4, 2, 6, 6, 4, 2),
  yend = c(7.45, 5.45, 5.45, 3.45, 3.45, 3.45)
)

# --- data inputs shown aside ---
data_labels <- data.frame(
  label = c("Z500 NetCDF\n(ERA5)", "Surface var. NetCDF\n(TX / precip)"),
  x = c(1.2, 7),
  y = c(9, 9)
)

ggplot() +
  # arrows
  geom_segment(data = arrows,
               aes(x = x, y = y, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
               linewidth = 0.7, colour = "grey40") +
  # data input arrows
  geom_segment(aes(x = 1.85, y = 9, xend = 2.8, yend = 9),
               arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
               colour = "#4575b4", linewidth = 0.6) +
  geom_segment(aes(x = 6.55, y = 9, xend = 5.8, yend = 7.1),
               arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
               colour = "#4575b4", linewidth = 0.6) +
  # node boxes
  geom_tile(data = nodes,
            aes(x = x, y = y, fill = type),
            width = 2.4, height = 1.4, colour = "white", linewidth = 1.2,
            alpha = 0.85) +
  # node labels
  geom_text(data = nodes,
            aes(x = x, y = y, label = label),
            size = 3, lineheight = 1.15, fontface = "bold", colour = "white") +
  # data labels
  geom_label(data = data_labels,
             aes(x = x, y = y, label = label),
             fill = "#e8f4fd", colour = "#2c5f8a", size = 2.8,
             label.padding = unit(0.35, "lines")) +
  # period annotation
  annotate("text", x = 4, y = 6.05, size = 2.8, colour = "grey40",
           label = "periods = c(1951,1980, 1993,2022)") +
  scale_fill_manual(
    values = pal,
    labels = c(input = "Data preparation", core = "Core methodology",
               output = "Attribution output", optional = "Optional (downscaling)"),
    name = NULL) +
  scale_x_continuous(limits = c(0.1, 8.2)) +
  scale_y_continuous(limits = c(2.1, 10.2)) +
  theme_void() +
  theme(legend.position  = "bottom",
        legend.text      = element_text(size = 9),
        plot.margin      = margin(10, 10, 10, 10))
```

---

## Setup

```{r load-pkg}
# Load functions directly from source (development mode).
# Replace with library(climattR) once the package is installed.
invisible(lapply(
  list.files(rprojroot::find_package_root_file("R"), full.names = TRUE),
  source
))
library(terra)
library(dplyr)
library(ggplot2)
```

---

## Step 1 — Prepare data {#step1}

`prepare_data()` processes a NetCDF raster (Z500), creates a calendar-based
temporal window around the event, and optionally detrends. It returns two
`SpatRaster` objects: the historical analogue pool and the event itself.

```{r prepare}
Z500 <- terra::rast(
  if (requireNamespace("climattR", quietly = TRUE))
    system.file("extdata", "z500_0509_1950_2023_eu.nc", package = "climattR")
  else
    rprojroot::find_package_root_file("inst", "extdata", "z500_0509_1950_2023_eu.nc")
)

event_dates <- seq(as.Date("2023-07-12"), as.Date("2023-07-18"), by = "day")

prep <- prepare_data(
  x           = Z500,
  event_dates = event_dates,
  time_window = 31,       # ±31 calendar days define the analog pool
  detrend     = FALSE
)

cat("Historical pool layers :", terra::nlyr(prep$ts_wo_event), "\n")
cat("Event layers (days)    :", terra::nlyr(prep$event),       "\n")
```

**Key parameters**

| Argument | Default | Purpose |
|----------|---------|---------|
| `time_window` | 31 | Days before/after event for the analog pool |
| `analog_months` | NULL | Month-based pool instead of time window |
| `detrend` | FALSE | Remove long-term polynomial trend |
| `k` | 2 | Polynomial degree for detrending |

---

## Step 2 — Find analogues {#step2}

`analogs_searcher()` computes a distance metric (RMSD by default) between every
day in the pool and each event day, then returns the top-$N$ matches per
sub-period.

```{r analogs}
analogs <- analogs_searcher(
  ts_wo_event = list(prep$ts_wo_event),
  event       = list(prep$event),
  n           = 20,
  periods     = c(1951, 1980, 1993, 2022),   # two climate sub-periods
  metric      = "rmsd"
)

cat("Full-period analogues  :", nrow(analogs$analogs_full_period), "\n")
cat("Sub-period analogues   :", nrow(analogs$analogs_subperiods),  "\n")
```

**Quality check** — RMSD should be comparable across periods:

```{r analog-quality, fig.height=3.5}
analogs$analogs_subperiods |>
  group_by(period, time_obj) |>
  summarise(mean_dist = mean(dist), .groups = "drop") |>
  ggplot(aes(x = time_obj, y = mean_dist, colour = period, group = period)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_colour_manual(values = c("1951-1980" = "#4575b4", "1993-2022" = "#d73027")) +
  labs(x = NULL, y = "Mean RMSD", colour = "Period",
       title = "Analog quality by event day and climate period") +
  theme_bw()
```

---

## Step 3a — Spatial bootstrap reconstruction {#step3a}

`bs_spatanalogs()` generates $n$ bootstrap replicates of the spatially
reconstructed surface field for each sub-period, expressed as anomalies with
respect to a reference climatology.

```{r bs-spat, fig.cap="Mean temperature anomaly (°C) for each climate sub-period. Attribution signal = factual − counterfactual.", fig.height=4, fig.width=10}
TX <- terra::rast(
  if (requireNamespace("climattR", quietly = TRUE))
    system.file("extdata", "tx_0608_1950_2023.nc", package = "climattR")
  else
    rprojroot::find_package_root_file("inst", "extdata", "tx_0608_1950_2023.nc")
)

bs_sp <- suppressMessages(bs_spatanalogs(
  x          = TX,
  analogs    = analogs$analogs_subperiods,
  n          = 50,          # use n=1000 for final results
  event_fun  = "mean",
  anom       = TRUE,
  ref_period = c(1950, 2022)
))

cf_mean  <- mean(bs_sp[["1951-1980"]])
fac_mean <- mean(bs_sp[["1993-2022"]])
attr_signal <- fac_mean - cf_mean

# Convert to ggplot2-friendly data frames
to_df <- function(r, label) {
  as.data.frame(r, xy = TRUE) |>
    dplyr::rename(value = 3) |>
    dplyr::mutate(panel = label)
}

plot_df <- dplyr::bind_rows(
  to_df(cf_mean,    "Counterfactual (1951-1980)"),
  to_df(fac_mean,   "Factual (1993-2022)"),
  to_df(attr_signal,"Attribution signal (factual \u2212 counterfactual)")
)

# shared colour limits: clip to 95th percentile to avoid outlier stretching
lim <- quantile(abs(plot_df$value), 0.95, na.rm = TRUE)

ggplot(plot_df, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  facet_wrap(~panel, nrow = 1) +
  scale_fill_gradient2(
    low      = "#4575b4",
    mid      = "white",
    high     = "#d73027",
    midpoint = 0,
    limits   = c(-lim, lim),
    oob      = scales::squish,
    name     = "\u00b0C"
  ) +
  coord_equal() +
  labs(x = NULL, y = NULL,
       title = "Bootstrap spatial reconstruction — 2023-07-12/18 heatwave") +
  theme_bw(base_size = 10) +
  theme(strip.text       = element_text(face = "bold"),
        axis.text        = element_blank(),
        axis.ticks       = element_blank(),
        panel.grid       = element_blank(),
        legend.position  = "right")
```

---

## Step 3b — Tabular bootstrap {#step3b}

For regional or station-scale analysis, `as_ts()` extracts a 1-D time series
from the raster and `bs_analogs()` runs the same bootstrap on the table.

```{r bs-tab}
TX <- terra::rast(
  if (requireNamespace("climattR", quietly = TRUE))
    system.file("extdata", "tx_0608_1950_2023.nc", package = "climattR")
  else
    rprojroot::find_package_root_file("inst", "extdata", "tx_0608_1950_2023.nc")
)

# Spatial mean over the full domain — supply 'domain' for a specific region
ts_data <- as_ts(TX)

bs_tab <- bs_analogs(
  x          = ts_data,
  analogs    = analogs$analogs_subperiods,
  n          = 500,
  anom       = TRUE,
  ref_period = c(1950, 2022)
)

# The result is a 'climattr_bs' S3 object
print(bs_tab)
```

---

## Step 4 — Attribution metrics {#step4}

`attribution_metrics()` computes the standard WWA metrics from the bootstrap
distributions:

$$
\text{PR} = \frac{P(\text{event} \mid \text{factual})}{P(\text{event} \mid \text{counterfactual})}
\qquad
\text{FAR} = 1 - \frac{1}{\text{PR}}
$$

```{r metrics}
# Use the observed event magnitude as threshold
obs_mean <- mean(bs_tab$observed$var, na.rm = TRUE)
cat("Observed event mean anomaly:", round(obs_mean, 1), "°C\n")

metrics <- attribution_metrics(
  x          = bs_tab,
  threshold  = obs_mean,
  tail       = "upper",   # hot event: P(X >= threshold)
  conf_level = 0.95,
  n_boot     = 1000
)

knitr::kable(metrics, digits = 3,
             caption = "Attribution metrics for the 2023-07-12/18 event")
```

**Interpretation guide**

| Metric | Value | Meaning |
|--------|-------|---------|
| PR > 1 | e.g. 5 | Event 5× more likely in current climate |
| FAR → 1 | e.g. 0.8 | 80% of probability attributable to CC |
| delta_T > 0 | e.g. +2°C | Distribution shifted +2°C by CC |

---

## Step 5 — Visualisation {#step5}

`autoplot()` produces a publication-ready density plot comparing the two
bootstrap distributions, with the observed value and the user-defined threshold.

```{r autoplot, fig.cap="Bootstrap distributions for the 2023 heatwave. Blue = counterfactual (1951–1980), red = factual (1993–2022). Dashed line = observed mean anomaly."}
autoplot(bs_tab, threshold = obs_mean,
         xlab  = "Temperature anomaly (°C)",
         title = "2023-07-12/18 heatwave — factual vs. counterfactual")
```

---

## Step 6 — Dynamic / thermodynamic decomposition {#step6}

`decompose_signal()` splits the total attribution signal into:

- **Dynamic**: driven by changes in atmospheric *circulation* between periods  
- **Thermodynamic**: driven by background *warming* at fixed circulation  

The method requires two `bs_analogs()` runs on the same analogs — one with raw
temperatures (`detrend = FALSE`) and one with detrended temperatures
(`detrend = TRUE`, which removes the long-term trend and isolates the
circulation signal).

$$
\Delta X_\text{total} = \underbrace{\Delta X_\text{dyn}}_{\text{circulation}}
                      + \underbrace{\Delta X_\text{thermo}}_{\text{warming}}
$$

```{r decompose, fig.cap="Decomposition of the 2023 heatwave signal into dynamic (circulation) and thermodynamic (warming) components. Error bars show 95% CI."}
bs_det <- suppressMessages(bs_analogs(
  ts_data, analogs$analogs_subperiods,
  n          = 500,
  anom       = TRUE,
  ref_period = c(1950, 2022),
  detrend    = TRUE   # <-- key difference
))

dcmp <- decompose_signal(bs_tab, bs_det)
knitr::kable(dcmp, digits = 3,
             caption = "Dynamic / thermodynamic decomposition (°C)")

plot_decomposition(dcmp,
                   title = "2023-07-12/18 heatwave — signal decomposition")
```

---

## Step 7 — EVT-based attribution {#step7}

`evt_metrics()` fits a parametric extreme value distribution (GEV or GPD) to
the bootstrap samples and derives return periods and PR/FAR from the fitted
model. This gives smoother probability estimates in the tails compared to
empirical counting.

```{r evt, fig.cap="Comparison of empirical (counting) vs. GEV-based PR and return periods."}
library(extRemes)

m_gev <- evt_metrics(bs_tab, threshold = obs_mean, type = "GEV", n_boot = 200)
m_gp  <- evt_metrics(bs_tab, threshold = obs_mean, type = "GP",  n_boot = 200)

knitr::kable(
  dplyr::bind_rows(
    dplyr::mutate(m_gev, method = "GEV"),
    dplyr::mutate(m_gp,  method = "GPD")
  ) |> dplyr::select(method, dist_type, return_period_cf, return_period_f,
                     PR, PR_low, PR_high, FAR),
  digits = 1,
  caption = "EVT-based attribution metrics. RP = return period (years)."
)
```

---

## Step 8 — Return level curves {#step8}

`return_level_curves()` fits a GEV to each climate sub-period separately and
produces the full return level-return period relationship for both worlds. The
vertical gap between the two curves at any given return period is the
**attribution signal** at that extremity — visually showing how the distribution
tail has shifted.

```{r return-levels, fig.cap="Return level curves for the counterfactual (1951–1980, blue) and factual (1993–2022, red) periods. Shaded bands are 95% CI. Dashed line marks the observed event anomaly.", fig.height=5, fig.width=7}
rl <- return_level_curves(bs_tab, n_boot = 300)

plot_return_levels(
  rl,
  obs_val = obs_mean,
  title   = "Return levels — 2023 heatwave factual vs. counterfactual"
)
```

The dashed line intersects the two curves at their respective **return periods
for the observed event intensity** — the shift between the two intersections is
the change in rarity attributable to climate change.

---

## Step 9 — Downscale (optional) {#step9}

`downscale_temperature()` refines the coarse reconstruction to high resolution
using kriging with elevation as covariate. Requires the `gstat` and `elevatr`
packages.

```{r downscale, eval=FALSE}
tx_hires <- downscale_temperature(
  x      = mean(bs_sp[["1993-2022"]]),
  analogs = analogs$analogs_subperiods,
  dem    = NULL,    # auto-downloaded via elevatr
  disagg = 4,       # 4× spatial refinement
  z      = 8        # DEM zoom level
)
terra::plot(tx_hires, main = "Downscaled TX anomaly (factual)")
```

---

## Full pipeline at a glance

```{r pipeline-summary, eval=FALSE}
library(climattR)
library(terra)

# ── Data ──────────────────────────────────────────────────────────────────────
Z500 <- rast("z500.nc")
TX   <- rast("tx.nc")
event_dates <- seq(as.Date("2023-07-12"), as.Date("2023-07-18"), by = "day")

# ── Step 1: Prepare ───────────────────────────────────────────────────────────
prep <- prepare_data(Z500, event_dates = event_dates, time_window = 31)

# ── Step 2: Analogues ─────────────────────────────────────────────────────────
analogs <- analogs_searcher(
  list(prep$ts_wo_event), list(prep$event),
  n = 20, periods = c(1951, 1980, 1993, 2022)
)

# ── Step 3: Bootstrap ─────────────────────────────────────────────────────────
ts_data <- as_ts(TX)
bs_tab  <- bs_analogs(ts_data, analogs$analogs_subperiods,
                      n = 1000, anom = TRUE, ref_period = c(1950, 2022))
bs_det  <- bs_analogs(ts_data, analogs$analogs_subperiods,
                      n = 1000, anom = TRUE, ref_period = c(1950, 2022),
                      detrend = TRUE)

# ── Step 4: Empirical metrics ─────────────────────────────────────────────────
obs_mean <- mean(bs_tab$observed$var)
metrics  <- attribution_metrics(bs_tab, threshold = obs_mean, tail = "upper")

# ── Step 5: EVT metrics ───────────────────────────────────────────────────────
metrics_gev <- evt_metrics(bs_tab, threshold = obs_mean, type = "GEV")

# ── Step 6: Decomposition ─────────────────────────────────────────────────────
dcmp <- decompose_signal(bs_tab, bs_det)

# ── Step 7: Visualise ─────────────────────────────────────────────────────────
autoplot(bs_tab, threshold = obs_mean)
plot_decomposition(dcmp)
```

---

## Session info

```{r session}
sessionInfo()
```
