---
title: "1. Heatwave attribution"
author: "Marc Lemus-Canovas"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{1. Heatwave attribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4
)
```

## Goal

Attribute the **July 2023 western European heatwave** to climate change using
flow analogues. The full pipeline runs in under 5 minutes on a laptop.

```{r load}
# Development mode: source functions directly
invisible(lapply(
  list.files(rprojroot::find_package_root_file("R"), full.names = TRUE), source
))
library(terra); library(dplyr); library(ggplot2); library(scales)
library(lubridate); library(stringr); library(tidyr)
library(pbapply); library(magrittr); library(tibble)

Z500 <- terra::rast(
  rprojroot::find_package_root_file("inst", "extdata", "z500_0509_1950_2023_eu.nc"))
TX <- terra::rast(
  rprojroot::find_package_root_file("inst", "extdata", "tx_0608_1950_2023.nc"))
```

## Step 1 — Define the event and build the analog pool

`prepare_data()` selects all days within ±31 calendar days of the event across
all years. These are the candidate analog days.

```{r prepare}
event_dates <- seq(as.Date("2023-07-12"), as.Date("2023-07-18"), by = "day")

prep <- prepare_data(
  x           = Z500,
  event_dates = event_dates,
  time_window = 31
)
cat("Analog pool:", terra::nlyr(prep$ts_wo_event), "days\n")
cat("Event:      ", terra::nlyr(prep$event), "days\n")
```

## Step 2 — Find the best analog days

`analogs_searcher()` ranks every day in the pool by RMSD to the event Z500
pattern and keeps the top 20 per sub-period.

```{r analogs}
analogs <- analogs_searcher(
  ts_wo_event = list(prep$ts_wo_event),
  event       = list(prep$event),
  n           = 20,
  periods     = c(1951, 1980,   # counterfactual world
                  1993, 2022),  # factual world
  metric      = "rmsd"
)
cat("Sub-period analogs:", nrow(analogs$analogs_subperiods), "\n")
```

## Step 3 — Bootstrap surface temperatures

For each bootstrap replicate, one analog day is drawn per event day. The
mean temperature anomaly across the multi-day event is computed.

```{r bootstrap}
ts_data <- as_ts(TX)   # convert raster → domain-mean time series

bs <- suppressMessages(
  bs_analogs(
    x          = ts_data,
    analogs    = analogs$analogs_subperiods,
    n          = 500,
    anom       = TRUE,
    ref_period = c(1950, 2022)
  )
)
print(bs)   # S3 summary
```

## Step 3b — Spatial bootstrap reconstruction

`bs_spatanalogs()` produces the same bootstrap but returns full gridded fields
instead of domain means. The mean over all replicates gives the expected
temperature anomaly map for each climate world.

```{r spatial-maps, fig.width=10, fig.height=4, fig.cap="Mean temperature anomaly (°C) reconstructed by flow analogues. Attribution signal = factual − counterfactual."}
bs_sp <- suppressMessages(
  bs_spatanalogs(
    x          = TX,
    analogs    = analogs$analogs_subperiods,
    n          = 50,          # use n=1000 for publication results
    anom       = TRUE,
    ref_period = c(1950, 2022)
  )
)

cf_mean  <- mean(bs_sp[["1951-1980"]])
fac_mean <- mean(bs_sp[["1993-2022"]])
signal   <- fac_mean - cf_mean

# Convert to ggplot2 data frames
to_df <- function(r, label) {
  as.data.frame(r, xy = TRUE) |>
    dplyr::rename(value = 3) |>
    dplyr::mutate(panel = label)
}

plot_df <- dplyr::bind_rows(
  to_df(cf_mean,  "Counterfactual (1951-1980)"),
  to_df(fac_mean, "Factual (1993-2022)"),
  to_df(signal,   "Attribution signal (F \u2212 CF)")
)

lim <- quantile(abs(plot_df$value), 0.95, na.rm = TRUE)

ggplot(plot_df, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  facet_wrap(~panel, nrow = 1) +
  scale_fill_gradient2(
    low = "#4575b4", mid = "white", high = "#d73027",
    midpoint = 0, limits = c(-lim, lim),
    oob = scales::squish, name = "\u00b0C"
  ) +
  coord_equal() +
  labs(x = NULL, y = NULL,
       title = "Bootstrap spatial reconstruction — July 2023 heatwave") +
  theme_bw(base_size = 10) +
  theme(strip.text = element_text(face = "bold"),
        axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank())
```

## Step 4 — Attribution metrics

The observed event magnitude is used as the threshold. PR and FAR compare
how often this threshold is exceeded in each climate world.

```{r metrics}
obs <- mean(bs$observed$var, na.rm = TRUE)
cat("Observed anomaly:", round(obs, 2), "°C\n")

m <- attribution_metrics(bs, threshold = obs, tail = "upper")
knitr::kable(m, digits = 2,
             caption = "Attribution metrics — July 2023 heatwave")
```

## Step 5 — Visualise the distributions

```{r autoplot, fig.cap="Blue = counterfactual (1951–1980), red = factual (1993–2022). Dashed line = observed anomaly."}
autoplot(bs,
         threshold = obs,
         xlab  = "Temperature anomaly (°C)",
         title = "July 2023 heatwave — attribution distributions")
```

**Reading the plot**: the factual distribution (red) is shifted to the right
relative to the counterfactual (blue). The observed event (dashed line) falls
deep in the tail of the counterfactual world but is common in the factual one.

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
