---
title: "2. Adapting to different events and variables"
author: "Marc Lemus-Canovas"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{2. Adapting to different events and variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4
)
```

## The same framework, any surface variable

The flow-analogue method is **variable-agnostic**: the circulation search
always uses Z500, but any gridded surface variable (temperature, precipitation,
wind speed, humidity) can be plugged in as the surface field.

| Variable | Typical tail | Notes |
|----------|-------------|-------|
| TX (max temperature) | upper | shown in vignette 1 |
| Precipitation | upper | divide by `scale` if stored as mm×10 |
| Minimum temperature | lower | set `tail = "lower"` in `attribution_metrics()` |
| Wind speed | upper | use daily max or mean |

The only adaptation needed is the **temporal window**: the Z500 and surface
NetCDF must overlap in time and cover the same season.

## Example — 2003 European heatwave

The August 2003 heatwave is one of the deadliest on record. We replicate the
attribution using the same TX dataset.

```{r load}
invisible(lapply(
  list.files(rprojroot::find_package_root_file("R"), full.names = TRUE), source
))
library(terra); library(dplyr); library(ggplot2)
library(lubridate); library(stringr); library(tidyr)
library(pbapply); library(magrittr); library(tibble)

Z500 <- terra::rast(
  rprojroot::find_package_root_file("inst", "extdata", "z500_0509_1950_2023_eu.nc"))
TX <- terra::rast(
  rprojroot::find_package_root_file("inst", "extdata", "tx_0608_1950_2023.nc"))
```

```{r event2003}
# 2003 European heatwave peak: 1–12 August 2003
event_2003 <- seq(as.Date("2003-08-01"), as.Date("2003-08-12"), by = "day")

prep <- prepare_data(Z500, event_dates = event_2003, time_window = 31)

analogs <- analogs_searcher(
  list(prep$ts_wo_event), list(prep$event),
  n = 20, periods = c(1951, 1980, 1993, 2022), metric = "rmsd"
)

ts_data <- as_ts(TX)
bs_2003 <- suppressMessages(
  bs_analogs(ts_data, analogs$analogs_subperiods,
             n = 500, anom = TRUE, ref_period = c(1950, 2022))
)
print(bs_2003)
```

```{r metrics2003}
obs_2003 <- mean(bs_2003$observed$var, na.rm = TRUE)
cat("Observed 2003 anomaly:", round(obs_2003, 2), "°C\n")

m_2003 <- attribution_metrics(bs_2003, threshold = obs_2003, tail = "upper")
knitr::kable(m_2003, digits = 2,
             caption = "Attribution metrics — August 2003 heatwave")
```

```{r plot2003, fig.cap="Distributions for the 2003 heatwave. Note that both distributions overlap more than 2023 — the 2003 event was partially explainable without climate change."}
autoplot(bs_2003, threshold = obs_2003,
         xlab  = "Temperature anomaly (°C)",
         title = "August 2003 heatwave — attribution distributions")
```

## Using precipitation data

For precipitation attribution the workflow is identical. The only things that
change are the surface NetCDF and the tail direction:

```{r precip, eval=FALSE}
# Load your precipitation raster (daily, mm or mm×10)
PR <- terra::rast("path/to/precip.nc")

# Convert to time series (domain mean)
ts_pr <- as_ts(PR)

# Bootstrap — same analogs, different surface
bs_pr <- bs_analogs(ts_pr, analogs$analogs_subperiods,
                    n = 500, anom = TRUE, ref_period = c(1950, 2022))

# For a wet extreme use tail = "upper"
obs_pr <- mean(bs_pr$observed$var, na.rm = TRUE)
attribution_metrics(bs_pr, threshold = obs_pr, tail = "upper")

# For a drought (dry extreme) use tail = "lower"
attribution_metrics(bs_pr, threshold = obs_pr, tail = "lower")
```

## Choosing the reference period

The `ref_period` in `bs_analogs()` sets the climatological baseline used to
compute anomalies:

- Use the **full data span** (e.g. `c(1950, 2022)`) for the broadest baseline.
- Use a **standard climatology** (e.g. `c(1981, 2010)`) for comparability with
  operational products.
- Set `anom = FALSE` to work in absolute units (useful for precipitation or wind).

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
