---
title: "3. Understanding analog metrics"
author: "Marc Lemus-Canovas"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{3. Understanding analog metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4
)
```

## What is an analog?

An **analog day** is a historical day whose large-scale circulation (Z500)
resembles the target event. The closer the Z500 pattern, the better the analog.

`analogs_searcher()` supports three distance metrics:

| Metric | Formula | Sensitive to |
|--------|---------|-------------|
| `rmsd` | $\sqrt{\frac{1}{n}\sum(z_i - e_i)^2}$ | Absolute differences |
| `euclidean` | $\sqrt{\sum(z_i - e_i)^2}$ | Absolute differences (not normalised) |
| `pearson` | $1 - r(z, e)$ | Pattern shape only, not amplitude |

```{r load}
invisible(lapply(
  list.files(rprojroot::find_package_root_file("R"), full.names = TRUE), source
))
library(terra); library(dplyr); library(ggplot2); library(tidyr)
library(lubridate); library(stringr); library(pbapply)
library(magrittr); library(tibble)

Z500 <- terra::rast(
  rprojroot::find_package_root_file("inst", "extdata", "z500_0509_1950_2023_eu.nc"))
TX <- terra::rast(
  rprojroot::find_package_root_file("inst", "extdata", "tx_0608_1950_2023.nc"))

event_dates <- seq(as.Date("2023-07-12"), as.Date("2023-07-18"), by = "day")
prep <- prepare_data(Z500, event_dates = event_dates, time_window = 31)
```

## Comparing the three metrics

```{r compare-metrics, fig.height=5, fig.cap="Distribution of analog RMSD distances per period and metric. Lower = better analogs."}
metrics_list <- lapply(c("rmsd", "euclidean", "pearson"), function(m) {
  a <- analogs_searcher(
    list(prep$ts_wo_event), list(prep$event),
    n = 20, periods = c(1951, 1980, 1993, 2022), metric = m
  )
  dplyr::mutate(a$analogs_subperiods, metric = m)
})

dplyr::bind_rows(metrics_list) |>
  ggplot(aes(x = metric, y = dist, fill = period)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.8) +
  scale_fill_manual(values = c("1951-1980" = "#4575b4", "1993-2022" = "#d73027")) +
  labs(x = "Metric", y = "Distance score", fill = "Period",
       title = "Analog quality by metric",
       subtitle = "Distances are not comparable across metrics (different scales)") +
  theme_bw()
```

> **Tip**: RMSD is the default and usually the best choice for Z500 fields.
> Pearson is useful when only the pattern shape matters (e.g. blocking index).

## Effect of n (number of analogs)

More analogs = wider bootstrap distribution = more uncertainty, but also less
bias from poor analogs. The sweet spot is usually n = 15–30.

```{r n-effect, fig.cap="Effect of n on the bootstrap variance. Shaded band = ±1 sd across bootstrap replicates."}
ts_data <- as_ts(TX)
analogs_base <- analogs_searcher(
  list(prep$ts_wo_event), list(prep$event),
  n = 40, periods = c(1951, 1980, 1993, 2022)
)

n_values <- c(5, 10, 20, 40)
results <- lapply(n_values, function(n_top) {
  # restrict to top-n per period/event day
  sub <- analogs_base$analogs_subperiods |>
    dplyr::group_by(period, time_obj) |>
    dplyr::slice_min(dist, n = n_top) |>
    dplyr::ungroup()
  bs <- suppressMessages(
    bs_analogs(ts_data, sub, n = 200, anom = TRUE, ref_period = c(1950, 2022))
  )
  bs$summary_bs |> dplyr::mutate(n_analogs = n_top)
})

dplyr::bind_rows(results) |>
  ggplot(aes(x = time_obj, y = var_mean,
             colour = period, group = interaction(period, n_analogs),
             linetype = factor(n_analogs))) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("1951-1980" = "#4575b4", "1993-2022" = "#d73027")) +
  scale_linetype_manual(values = c("5"="dotted","10"="dashed","20"="solid","40"="solid"),
                        name = "n analogs") +
  labs(x = NULL, y = "Mean anomaly (°C)", colour = "Period",
       title = "Effect of n on bootstrap mean") +
  theme_bw()
```

## Analog quality diagnostic

A good attribution requires similar analog quality in both periods — if RMSD is
much higher in one period, the analog pool is unbalanced.

```{r quality, fig.cap="Mean RMSD per event day. Both periods should be comparable."}
analogs_base$analogs_subperiods |>
  dplyr::group_by(period, time_obj) |>
  dplyr::summarise(mean_rmsd = mean(dist), .groups = "drop") |>
  ggplot(aes(x = time_obj, y = mean_rmsd, colour = period, group = period)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c("1951-1980" = "#4575b4", "1993-2022" = "#d73027")) +
  labs(x = "Event day", y = "Mean RMSD", colour = "Period",
       title = "Analog quality check — RMSD should be comparable across periods") +
  theme_bw()
```

If one period systematically shows higher RMSD, try:

- Increasing `n` to allow worse analogs in the sparse period
- Widening `time_window` to expand the seasonal pool
- Checking that the data covers the full period requested

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
