---
title: "4. Advanced attribution: decomposition and return levels"
author: "Marc Lemus-Canovas"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{4. Advanced attribution: decomposition and return levels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.5
)
```

## Prerequisites

This vignette builds on the basics from vignette 1. We start from the
bootstrap output and explore advanced analyses.

```{r load}
invisible(lapply(
  list.files(rprojroot::find_package_root_file("R"), full.names = TRUE), source
))
library(terra); library(dplyr); library(ggplot2); library(extRemes)
library(lubridate); library(stringr); library(tidyr); library(scales)
library(pbapply); library(magrittr); library(tibble); library(pracma)

Z500 <- terra::rast(
  rprojroot::find_package_root_file("inst", "extdata", "z500_0509_1950_2023_eu.nc"))
TX <- terra::rast(
  rprojroot::find_package_root_file("inst", "extdata", "tx_0608_1950_2023.nc"))

event_dates <- seq(as.Date("2023-07-12"), as.Date("2023-07-18"), by = "day")
prep    <- prepare_data(Z500, event_dates = event_dates, time_window = 31)
analogs <- analogs_searcher(
  list(prep$ts_wo_event), list(prep$event),
  n = 20, periods = c(1951, 1980, 1993, 2022)
)
ts_data <- as_ts(TX)

# Base bootstrap (raw temperatures)
bs_raw <- suppressMessages(
  bs_analogs(ts_data, analogs$analogs_subperiods,
             n = 500, anom = TRUE, ref_period = c(1950, 2022), detrend = FALSE)
)
obs <- mean(bs_raw$observed$var, na.rm = TRUE)
cat("Observed anomaly:", round(obs, 2), "°C\n")
```

## Part 1 — Dynamic / thermodynamic decomposition

The total attribution signal has two physical causes:

- **Dynamic**: changes in atmospheric *circulation* between periods
- **Thermodynamic**: direct background *warming* at fixed circulation

We run a second bootstrap with **detrended** temperatures (`detrend = TRUE`).
Detrending removes the long-term warming trend while keeping the day-to-day
circulation variability. The dynamic component is then the difference
between factual and counterfactual in the detrended world.

$$\Delta T_\text{total} = \Delta T_\text{dyn} + \Delta T_\text{thermo}$$

```{r decompose-run}
# Detrended bootstrap (circulation signal only)
bs_det <- suppressMessages(
  bs_analogs(ts_data, analogs$analogs_subperiods,
             n = 500, anom = TRUE, ref_period = c(1950, 2022), detrend = TRUE)
)

dcmp <- decompose_signal(bs_raw, bs_det)
knitr::kable(dcmp, digits = 2,
             caption = "Signal decomposition (°C). CI = 95% bootstrap confidence interval.")
```

```{r decompose-plot, fig.cap="The thermodynamic (warming) component clearly dominates the attribution signal for the 2023 heatwave. The slightly negative dynamic component means the 2023 circulation was marginally less extreme than average."}
plot_decomposition(
  dcmp,
  title = "July 2023 heatwave — dynamic / thermodynamic decomposition"
)
```

### Spatial decomposition maps

The decomposition can also be visualised spatially using `bs_spatanalogs()`
with both raw and detrended temperatures. Each pixel shows its own
dynamic / thermodynamic breakdown.

```{r spatial-decomp, fig.width=10, fig.height=8, fig.cap="Spatial decomposition of the attribution signal. Top row: raw bootstrap means (CF and factual). Bottom row: dynamic and thermodynamic components pixel by pixel."}
# Spatial bootstrap — raw
bs_sp_raw <- suppressMessages(
  bs_spatanalogs(TX, analogs$analogs_subperiods, n = 50,
                 anom = TRUE, ref_period = c(1950, 2022))
)
# Spatial bootstrap — detrended (removes long-term trend pixel by pixel)
bs_sp_det <- suppressMessages(
  bs_spatanalogs(TX, analogs$analogs_subperiods, n = 50,
                 anom = TRUE, ref_period = c(1950, 2022),
                 detrend = TRUE)   # <-- key difference
)

cf_raw  <- mean(bs_sp_raw[["1951-1980"]])
fac_raw <- mean(bs_sp_raw[["1993-2022"]])
total   <- fac_raw - cf_raw

cf_det  <- mean(bs_sp_det[["1951-1980"]])
fac_det <- mean(bs_sp_det[["1993-2022"]])
dyn     <- fac_det - cf_det
thermo  <- total - dyn

to_df <- function(r, label) {
  as.data.frame(r, xy = TRUE) |>
    dplyr::rename(value = 3) |>
    dplyr::mutate(panel = label)
}

plot_df <- dplyr::bind_rows(
  to_df(cf_raw,  "CF mean (1951-1980)"),
  to_df(fac_raw, "Factual mean (1993-2022)"),
  to_df(dyn,     "Dynamic component"),
  to_df(thermo,  "Thermodynamic component")
)
plot_df$panel <- factor(plot_df$panel,
  levels = c("CF mean (1951-1980)", "Factual mean (1993-2022)",
             "Dynamic component",   "Thermodynamic component"))

lim <- quantile(abs(plot_df$value), 0.95, na.rm = TRUE)

ggplot(plot_df, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  facet_wrap(~panel, nrow = 2, ncol = 2) +
  scale_fill_gradient2(
    low = "#4575b4", mid = "white", high = "#d73027",
    midpoint = 0, limits = c(-lim, lim),
    oob = scales::squish, name = "\u00b0C"
  ) +
  coord_equal() +
  labs(x = NULL, y = NULL,
       title = "Spatial decomposition \u2014 July 2023 heatwave") +
  theme_bw(base_size = 10) +
  theme(strip.text = element_text(face = "bold"),
        axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank())
```

## Part 2 — Return level curves

Return level curves show **how the full tail of the distribution has shifted**
between the two climate worlds. A GEV distribution is fitted to each period's
bootstrap samples separately.

```{r rl, fig.cap="Return level curves with 95% CI bands. The vertical gap between the two curves at any return period is the attribution signal at that extremity. The dashed line shows the observed event."}
rl <- return_level_curves(bs_raw, n_boot = 300)

plot_return_levels(
  rl,
  obs_val = obs,
  title   = "Return levels — July 2023 heatwave"
)
```

Reading the plot: the **horizontal distance** between the observed line and each
curve gives the return period of the event in that world. The **vertical
distance** between the two curves at a fixed return period gives the shift in
temperature anomaly (°C).

## Part 3 — EVT-based attribution metrics

Instead of counting exceedances empirically, `evt_metrics()` evaluates the
probability of exceeding the observed threshold from the fitted EVT model.
This gives smoother estimates, especially deep in the tails.

```{r evt}
m_gev <- evt_metrics(bs_raw, threshold = obs, type = "GEV", n_boot = 200)
m_gpd <- evt_metrics(bs_raw, threshold = obs, type = "GP",  n_boot = 200)

knitr::kable(
  dplyr::bind_rows(
    dplyr::mutate(m_gev, method = "GEV"),
    dplyr::mutate(m_gpd, method = "GPD")
  ) |> dplyr::select(method, return_period_cf, return_period_f,
                     PR, PR_low, PR_high, FAR),
  digits = 1,
  caption = "EVT attribution metrics. RP = return period (number of bootstrap samples)."
)
```

> **GEV vs. GPD**: GEV extrapolates further into the tail but with much wider
> CI. GPD is more conservative — it only models values above the 75th
> percentile threshold. For most attribution studies, GPD gives more honest
> uncertainty bounds when sample size is moderate (n = 500–1000).

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
